\Annexes


\Annex{List of Examples}
\label{chap:list-of-examples}

\begin{exe}
  \exr{ex:intro} She might still be dating that idiot.
  \dotfill \pageref{ex:intro}
\end{exe}

\begin{exe}
  \exr{ex:syllogism-hypothesis} Every man is mortal. Socrates is a man.
  \dotfill \pageref{ex:syllogism-hypothesis}
\end{exe}

\begin{exe}
  \exr{ex:syllogism-conclusion} Socrates is mortal.
  \dotfill \pageref{ex:syllogism-conclusion}
\end{exe}

\begin{exe}
  \exr{ex:no-antecedent} It fascinates him.
  \dotfill \pageref{ex:no-antecedent}
\end{exe}

\begin{exe}
  \exr{ex:ulysses} Jones$_1$ owns Ulysses$_2$. It$_2$ fascinates him$_1$.
  \dotfill \pageref{ex:ulysses}
\end{exe}

\begin{exe}
  \exr{ex:jones-porsche} Jones$_1$ owns a Porsche$_2$. It$_2$ fascinates him$_1$.
  \dotfill \pageref{ex:jones-porsche}
\end{exe}

\begin{exe}
  \exr{ex:donkey-rc} Every farmer who owns a donkey$_1$ beats it$_1$.
  \dotfill \pageref{ex:donkey-rc}
\end{exe}

\begin{exe}
  \exr{ex:donkey-if} If a farmer$_1$ owns a donkey$_2$, he$_1$ beats it$_2$.
  \dotfill \pageref{ex:donkey-if}
\end{exe}

\begin{exe}
  \exr{ex:donkey-cruel}[*]{Every farmer$_1$ who owns a donkey$_2$ beats
    it$_2$. He$_1$ is cruel.
  \dotfill \pageref{ex:donkey-cruel}}
\end{exe}

\begin{exe}
  \exr{ex:trivial} John loves Mary.
  \dotfill \pageref{ex:trivial}
\end{exe}

\begin{exe}
  \exr{ex:deixis} John loves me.
  \dotfill \pageref{ex:deixis}
\end{exe}

\begin{exe}
  \exr{ex:indirect-speech} John said Mary loves me.
  \dotfill \pageref{ex:indirect-speech}
\end{exe}

\begin{exe}
  \exr{ex:direct-speech} John said ``Mary loves me''.
  \dotfill \pageref{ex:direct-speech}
\end{exe}

\begin{exe}
  \exr{ex:either-or} Either John loves Sarah, or Mary, John's best friend, loves John.
  \dotfill \pageref{ex:either-or}
\end{exe}

\begin{exe}
  \exr{ex:if-not} If it is not the case that John, whom Sarah loves, loves Sarah then Mary loves John.
  \dotfill \pageref{ex:if-not}
\end{exe}

\begin{exe}
  \exr{ex:quantifiers} Every man loves a woman.
  \dotfill \pageref{ex:quantifiers}
\end{exe}

\begin{exe}
  \exr{ex:siamese} Every owner of a siamese cat loves a therapist.
  \dotfill \pageref{ex:siamese}
\end{exe}

\begin{exe}
  \exr{ex:mixed-nps} John loves a man.
  \dotfill \pageref{ex:mixed-nps}
\end{exe}

\begin{exe}
  \exr{ex:man-porsche} A man owns a Porsche. It fascinates him.
  \dotfill \pageref{ex:man-porsche}
\end{exe}

\begin{exe}
  \exr{ex:jones-mercedes} It is not the case that Jones$_1$ owns a
  Porsche. He$_1$ owns a Mercedes.
  \dotfill \pageref{ex:jones-mercedes}
\end{exe}

\begin{exe}
  \exr{ex:not-john-car} It is not the case that John$_1$ likes his$_1$ car.
  \dotfill \pageref{ex:not-john-car}
\end{exe}

\begin{exe}
  \exr{ex:cancel} If John owns a car, then his car is cheap.
  \dotfill \pageref{ex:cancel}
\end{exe}

\begin{exe}
  \exr{ex:john-presupposed} If there is a poor man called John, then if
  John owns a car, his car is cheap.
  \dotfill \pageref{ex:john-presupposed}
\end{exe}

\begin{exe}
  \exr{ex:wilma} ($c_0$) Maybe ($c_1$) Wilma thinks that ($c_2$) her
  husband is having an affair.
  \dotfill \pageref{ex:wilma}
\end{exe}

\begin{exe}
  \exr{ex:most-germans-wash} Most Germans wash their car on Saturday.
  \dotfill \pageref{ex:most-germans-wash}
\end{exe}

\begin{exe}
  \exr{ex:man-angry} ($c_0$) If ($c_1$) a man gets angry, ($c_2$) his
  children get frightened.
  \dotfill \pageref{ex:man-angry}
\end{exe}

\begin{exe}
  \exr{ex:cataphora}[*]{He$_1$ loves John's$_1$ car.
  \dotfill \pageref{ex:cataphora}}
\end{exe}

\begin{exe}
  \exr{ex:bathroom} Either there's no bathroom$_1$ in the house, or
  it's$_1$ in a funny place.
  \dotfill \pageref{ex:bathroom}
\end{exe}

\begin{exe}
  \exr{ex:hospital} John, who nearly killed a woman$_1$ with his car,
  visited her$_1$ in the hospital.
  \dotfill \pageref{ex:hospital}
\end{exe}

\begin{exe}
  \exr{ex:porridge} The porridge is warm.
  \dotfill \pageref{ex:porridge}
\end{exe}

\begin{exe}
  \exr{ex:crossover}[*]{His$_1$ mother likes every man$_1$.
  \dotfill \pageref{ex:crossover}}
\end{exe}

\begin{exe}
  \exr{ex:inverse} A woman loves every man.
  \dotfill \pageref{ex:inverse}
\end{exe}

\begin{exe}
  \exr{ex:primary-crossover}[*]{He$_1$ likes every man$_1$.
  \dotfill \pageref{ex:primary-crossover}}
\end{exe}

\begin{exe}
  \exr{ex:secondary-crossover}[*]{He$_1$ likes every man$_1$'s mother.
  \dotfill \pageref{ex:secondary-crossover}}
\end{exe}

\begin{exe}
  \exr{ex:bad-crossover}[*]{It$_1$ loves every owner of a dog$_1$.
  \dotfill \pageref{ex:bad-crossover}}
\end{exe}

\begin{exe}
  \exr{ex:presupposition-binding} The man who owns a dog$_1$ loves it$_1$.
  \dotfill \pageref{ex:presupposition-binding}
\end{exe}

\begin{exe}
  \exr{ex:presupposition-binding2} The man$_1$ who loves his$_1$ dog$_2$
  treats it$_2$ well.
  \dotfill \pageref{ex:presupposition-binding2}
\end{exe}

\begin{exe}
  \exr{ex:presupposition-every-narrow} The owner of every book ever
  published by Elsevier must be very rich.
  \dotfill \pageref{ex:presupposition-every-narrow}
\end{exe}

\begin{exe}
  \exr{ex:presupposition-every-wide} Mary sent an email to the
  representative of every country.
  \dotfill \pageref{ex:presupposition-every-wide}
\end{exe}

\begin{exe}
  \exr{ex:every-presupposition} Everyone$_1$ hates the sound of their$_1$
  voice.
  \dotfill \pageref{ex:every-presupposition}
\end{exe}

\begin{exe}
  \exr{ex:binding-in-presupposition} John met the woman who stole a car$_1$
  by hacking into its$_1$ computer.
  \dotfill \pageref{ex:binding-in-presupposition}
\end{exe}

\begin{exe}
  \exr{ex:combo} My best friend, who owns a dog$_1$, said it$_1$ loves everyone.
  \dotfill \pageref{ex:combo}
\end{exe}

\begin{exe}
  \exr{ex:C-operator} A man$_1$ walks in the park. He$_1$ whistles.
  \dotfill \pageref{ex:C-operator}
\end{exe}

\begin{exe}
  \exr{ex:mary-peter} Peter said ``Mary kissed me''.
  \dotfill \pageref{ex:mary-peter}
\end{exe}

\begin{exe}
  \exr{ex:peter-mary} Peter said ``I was kissed by Mary''.
  \dotfill \pageref{ex:peter-mary}
\end{exe}

\begin{exe}
  \exr{ex:modal-subordination} A wolf$_1$ might walk in. It$_1$ would growl.
  \dotfill \pageref{ex:modal-subordination}
\end{exe}

\begin{exe}
  \exr{ex:salmon}[*]{John had a great evening last night. He had a great
    meal. He ate salmon$_1$. He devoured lots of cheese. He then won a
    dancing competition. It$_1$ was pink.
  \dotfill \pageref{ex:salmon}}
\end{exe}
